---
title: "Hunting Model"
author: "Amanda Overbye and Matteo Torres"
date: "June 3, 2025"
format: 
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

## Load Packages

```{r, message = FALSE}
library(here)
library(purrr)
library(deSolve)
library(tidyverse)
```

## Source Functions

```{r}
source(here("lotcvhunt.R"))
```

## Model Parameters and Setup

```{r}
# parameter values
base_pars <- list(
  rprey = 0.95,
  alpha = 0.01,
  eff = 0.6,
  pmort = 0.4,
  K = 2000
)
# initial populations
init_pop <- c(prey = 100, pred = 10)
# time steps (100 years)
times <- seq(0, 100, by = 1)
```

## Defining Stability


## Stability Assessment Function


```{r}
check_stability <- function(rhunt, prey_min) {
  pars <- c(base_pars, rhunt = rhunt, prey_min = prey_min)
  sim <- ode(func = lotvhunt, y = init_pop, times = times, parms = pars)
  
  final_prey <- tail(sim[, "prey"], 1)
  final_pred <- tail(sim[, "pred"], 1)
  
  return(final_prey >= 50 & final_pred >= 10
)
}
```

## Parameter Space Exploration

```{r, warning = FALSE}
# define parameter grid
rhunt_vals <- seq(0, 0.05, by = 0.005)
prey_min_vals <- seq(100, 1000, by = 100)
param_grid <- expand.grid(rhunt = rhunt_vals, prey_min = prey_min_vals)

# run model for all combinations
stability_results <- param_grid %>%
  mutate(stable = pmap_lgl(list(rhunt, prey_min), function(rh, pm) {
    check_stability(rh, pm)
  }))
```

## Stability Analysis Results

```{r}
# Show stable parameter combinations
stable_combos <- stability_results %>% filter(stable == TRUE)
cat("Number of stable parameter combinations:", nrow(stable_combos), "out of", nrow(param_grid), "\n")

# Summary of stable hunting rates
stable_summary <- stable_combos %>%
  group_by(rhunt) %>%
  summarise(
    min_prey_min = min(prey_min),
    max_prey_min = max(prey_min),
    count = n(),
    .groups = 'drop'
  )

print("Stable hunting rates and corresponding minimum prey thresholds:")
stable_summary
```

## Visualization of Stability Landscape

```{r}
ggplot(stability_results, aes(x = rhunt, y = prey_min, fill = stable)) +
  geom_tile() +
  scale_fill_manual(values = c("FALSE" = "firebrick", "TRUE" = "darkblue"), 
                    name = "Stable") +
  labs(
    title = "Stability Landscape for Hunting Parameters",
    x = "Hunting Rate (rhunt)",
    y = "Minimum Prey Population Before Hunting",
    caption = "Green = Stable, Red = Unstable"
  ) +
  theme_minimal()
```

## Example Scenarios

```{r}
scenarios <- tibble(
  name = c("Low", "Moderate", "High", "Extreme"),
  rhunt = c(0.005, 0.015, 0.03, 0.05),
  prey_min = c(100, 300, 500, 700)
)

scenario_results <- scenarios %>%
  mutate(
    output = map2(rhunt, prey_min, ~ {
      pars <- c(base_pars, rhunt = .x, prey_min = .y)
      as.data.frame(ode(y = init_pop, times = times, func = lotvhunt, parms = pars))
    }),
    stable = map2_lgl(rhunt, prey_min, check_stability)
  )

# Show which scenarios are stable
scenario_stability <- scenarios %>%
  mutate(stable = map2_lgl(rhunt, prey_min, check_stability))
print("Scenario stability:")
scenario_stability
```

## Population Dynamics Visualization

```{r, warning = FALSE}
results_long <- scenario_results %>%
  unnest(output) %>%
  pivot_longer(cols = c(prey, pred), names_to = "species", values_to = "population")

ggplot(results_long, aes(x = time, y = population, color = species)) +
  geom_line(size = 1) +
  facet_wrap(~ name, ncol = 2) +
  labs(
    title = "Population Dynamics Under Different Hunting Scenarios",
    x = "Time (years)",
    y = "Population Size",
    color = "Species"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Conclusion

After running our analysis and testing various hunting rates, we concluded that a hunting rate of approximately 0.015 is ideal for maintaining stable predator and prey population sizes in the Lotka-Volterra model. We selected this rate based on our stability analysis, which visually demonstrated how different hunting rates influence population dynamics, particularly in relation to minimum prey thresholds required for hunting to occur. This conclusion was further supported by simulations of predefined hunting scenarios, where we compared population trajectories over time. As shown in the final visualization, the moderate hunting rate of 0.015 allows for sustainable prey removal without causing population collapse, thereby balancing ecological stability with hunting demands.